Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-05-09 manuscript/revision2/revision2.tex
manuscript: add response to reviews folder within manuscript/revision2/ and move intermediate files to intermediate_files folder

\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\firstpage[15]{2}

@@

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, Gulf vegetative growth
occurred before Midwestern vegetative growth, and Gulf flowering
occurred after Midwestern flowering (Figure~\ref{fig-map} A). At the
four northernmost common gardens (hereafter `North' gardens), located
within the natural range of the Midwest subpopulation, both Gulf
vegetative growth and flowering occurred after Midwest vegetative growth
and flowering. At the Oklahoma common garden, located near the natural
range limits of both the Gulf and the Midwest subpopulations, Gulf and
Midwest vegetative growth occurred over the same period, and Gulf
flowering occurred after Midwestern flowering (Figure~\ref{fig-map} A).
These patterns led to strong negative phenotypic correlations for the
timing of vegetative growth between the North and Texas gardens,
particularly in the Gulf and across all individuals (hereafter, `Both'
subpopulations), and contributed to positive phenotypic correlations for
the timing of flowering that had larger magnitudes at more northern
gardens (Figure~\ref{fig-map} B).

@@

We next looked for evidence of GxWeather by jointly re-estimating
genetic effects across all eight common gardens using the mash model,
which can flexibly capture many modes of covariance across gardens
(Figure~\ref{fig-map} D, SI Appendix, Datasets 1-6). We specified three
qualitative categories of covariance structures across gardens:
``canonical'' covariance, with simple patterns of effect size covariance
introduced in the initial \emph{mash} manuscript; ``data-driven''
covariances derived from common patterns of SNP effects observed in the
data, and ``GxWeather'' covariance, estimated from the covariance of
empirical weather patterns at each garden at specific times before the
phenological event (Table 1, SI Appendix, Section S2). For example, the
more similar that the amount of rainfall prior to flowering is for
genotypes A and B between gardens in MI and MO, the nearer these
covariances will be to one. If rainfalls are uncorrelated, covariance
will be near zero; if there is rank changing (e.g.~high rainfall before
flowering of A at MO but not MI, and high rainfall before flowering for
B at MI but not MO), this covariance will be near negative one. We
specified multiple covariance structures for the canonical, data-driven,
and GxWeather categories, each of which represented a specific pattern
of covariance in genetic effect size estimates that could be included in
the model. The phenotypic correlations for the timing of vegetative
growth had moderate negative correlations between the Texas and North
gardens, particularly in the Gulf subpopulation and Both subpopulations
(Figure~\ref{fig-map} B). If these phenotypic correlations have a
genetic component, they could be partially or completely captured by the
covariance structures specified in the mash model.

@@

Five GxWeather covariance structures were selected by the greedy
algorithm, zero to two per subpopulation: two for the start of
vegetative growth, and three for flowering date. Of these five, three
had mass on them in mash models of the strong effects
(Figure~\ref{fig-covar} A-B). Two of these three matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while one had all positive or near-zero
covariances. SNP-associated phenotypic effects covaried with different
weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the mash model of Gulf vegetative growth fell on a
covariance matrix constructed using the covariance of daylength 14 days
prior to the date of vegetative growth. The covariance matrix for this
weather cue was very similar to the pattern of phenotypic correlation
for the timing of vegetative growth in the Gulf subpopulation
(Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Mash models of the
timing of vegetative growth in the Midwest subpopulation and Both
subpopulations did not include this GxWeather covariance; the Midwest
had no weight on any GxWeather covariance, while Both subpopulations had
non-zero weights on two additional GxWeather covariance types, average
temperature one day prior to green-up, and the day length change in
seconds in the day prior to green-up (Figure~\ref{fig-covar} B). The
average temperature covariance matrix had negative covariances between
Texas and North gardens, though not as strong as the negative phenotypic
correlations seen in Both subpopulations (Figure~\ref{fig-map} B;
Figure~\ref{fig-covar} A). Only the Gulf subpopulation and Both
subpopulations had mass on any GxWeather covariance matrices
(Figure~\ref{fig-covar} C).

For flowering date, distinct GxWeather covariance structures captured
covariance in effect sizes for SNPs in the Gulf and Midwest
subpopulations. 33\% of SNP effects on flowering in the Gulf
subpopulation covaried with cumulative rainfall in the seven days prior
to flowering (Figure~\ref{fig-covar} D). 22.6\% of SNP effects on
flowering in the Midwest subpopulation covaried with day length change
in the two days prior to flowering (Figure~\ref{fig-covar} D), which had
negative covariances between Texas and North gardens. Neither covariance
matrix was selected as significantly improving the log-likelihood of the
\emph{mash} model of Both subpopulations, and no GxWeather covariance
matrices had a non-zero mass in models of Both subpopulations
(Figure~\ref{fig-covar} E). In five of the six mash models of strong
effects, the GxWeather covariance matrices captured a minority of the
posterior weights of the strong effects (Figure~\ref{fig-covar} C,E);
the majority of this mass was on various canonical covariance matrices.
These matrices included simple heterozygosity, with intermediate,
positive covariances between all gardens, and garden-specific effects
present only at one garden.

We next characterized the pairwise patterns of effects where we were
confident in the sign of the effect at both gardens. If the overall
pattern of phenotypic expression differs between a pair of gardens, this
difference may be comprised of SNP effects that differ in sign between
these gardens (effects with antagonistic pleiotropy), SNP effects that
differ in magnitude - that are large in one garden or region and smaller
in others, and SNP effects that are indistinguishable in the two gardens
(Figure~\ref{fig-effects} A). We used the local false sign rate (lfsr),
an analogue of the lfdr that establishes confidence in the effect sign,
not the effect's difference from zero, to determine significance. We
required lfsr significance (p \textless{} 0.05) in both gardens to
include effects. This means that our tests for antagonistic pleiotropy,
or a sign change between conditions, carry an equal statistical burden
to those for effects with the same sign.

@@

Identifying the environmental cues that are predictive of, or even
correlated with, plant phenotypic responses remains a major challenge to
studies interrogating gene action across many natural environments. The
GxWeather photoperiod and cumulative rainfall cues we identify here are
functions of the genotypes measured and capture only a minority of SNP
effects on flowering. We know still less about the overwintering
parameters that cause variation in the start of vegetative growth. We
could only assign SNP effects to a GxWeather covariance structures in
four of the six phenotype \& genetic subpopulations we modeled. More
generally, it is difficult to predict the time scales over which
individuals may integrate environmental cues, particularly in perennial
species which may integrate these cues over longer time scales. If this
integration time itself varies between individuals, we cannot select a
covariance structure that reflects this, though these structures would
likely be highly correlated with GxWeather structures we did include.
Our approach offers an opportunity to specify multiple environmental
cues and compete them to explain patterns of genetic effects, allowing
us to detect how important these cues are genome-wide, and how strongly
each cue influences each SNP. This is a key development to further
improve our understanding of genetic variation in GxE.

@@

To evaluate the prevalence and kinds of covariance patterns of SNP
effects across our common gardens, we used multivariate adaptive
shrinkage (\emph{mash}) on SNP effect estimates from the diversity panel
(35). To do this, we first conducted univarate genome-wide association
(GWAS) on site-specific best linear unbiased predictors (BLUPs) for the
timing of vegetative growth and flowering, using the switchgrassGWAS
package and the methods in (40). We used the effect estimates for single
nucleotide polymorphisms (SNPs) from these GWAS as preliminary effect
estimates for both traits at each garden. We then jointly modeled
genetic effect estimates, and the mixture of ways these effects might
covary, across all eight common gardens, using a two-part procedure. We
first modeled genetic effects with different combinations of covariance
structures, using the effect size estimates from a subset of \ul{19K}
relatively unlinked (r\textsuperscript{2} \textless{} 0.2), randomly
selected SNPs used in the GWAS. In these models (hereafter referred to
as ``random''), we then specified a variety of potential covariance
structures among genetic effect size estimates from each garden (as
expanded on below) and used a greedy algorithm to iteratively select and
add covariance structures that significantly improved the log likelihood
of the \emph{mash} model. Given the high correlation between some of the
covariance patterns we tested, this approach allowed us to select only
the subset of patterns of covariance that significantly improved the fit
of the jointly inferred structure of genetic effect estimates present in
the data. We then selected 19K relatively unlinked (r\textsuperscript{2}
\textless{} 0.2) SNPs from each univariate GWAS that had the largest
effect estimates in each GWAS (hereafter ``strong'' effects), which gave
sets of \ul{\textbf{48K - 80K}} SNPs, and jointly re-estimated these
effect size estimates at all eight sites using the set of significant
covariance matrices that most improved the model fit on the random
effects. Notably, the majority of these ``strong'' effects were not
significant in univariate GWAS.

>>>>

\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\firstpage[17]{2}

@@

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, Gulf vegetative growth
occurred before Midwestern vegetative growth, and Gulf flowering
occurred after Midwestern flowering (Figure~\ref{fig-map} A). At the
four northernmost common gardens (hereafter `North' gardens), located
within the natural range of the Midwest subpopulation, both Gulf
vegetative growth and flowering occurred after Midwest vegetative growth
and flowering. At the Oklahoma common garden, located near the natural
range limits of both the Gulf and the Midwest subpopulations, Gulf and
Midwest vegetative growth occurred over the same period, and Gulf
flowering occurred after Midwestern flowering (Figure~\ref{fig-map} A).
These patterns led to strong negative phenotypic correlations for the
timing of the start of vegetative growth between the North and Texas
gardens, particularly in the Gulf and across all individuals from the
Gulf and Midwest subpopulations (hereafter, `Both' subpopulations), and
contributed to positive phenotypic correlations for the timing of
flowering that had larger magnitudes at more northern gardens
(Figure~\ref{fig-map} B).

@@

We next looked for evidence of GxWeather by jointly re-estimating
genetic effects across all eight common gardens using the mash model,
which can flexibly capture many modes of covariance across gardens
(Figure~\ref{fig-map} D, SI Appendix, Datasets 1-6). We specified three
qualitative categories of covariance structures across gardens:
``canonical'' covariance, with simple patterns of effect size covariance
introduced in the initial \emph{mash} manuscript; ``data-driven''
covariances derived from common patterns of SNP effects observed in the
data, and ``GxWeather'' covariance, estimated from the covariance of
empirical weather patterns at each garden at specific times before the
phenological event (Table 1, SI Appendix, Section S2). For example, the
more similar that the amount of rainfall prior to flowering is for
genotypes A and B between gardens in MI and MO, the nearer these
covariances will be to one. If rainfall amounts are uncorrelated,
covariance will be near zero; if there is rank changing (e.g.~high
rainfall before flowering of A at MO but not MI, and high rainfall
before flowering for B at MI but not MO), this covariance will be near
negative one. We specified multiple covariance structures for the
canonical, data-driven, and GxWeather categories, each of which
represented a specific pattern of covariance in genetic effect size
estimates that could be included in the model. The phenotypic
correlations for the timing of vegetative growth had moderate negative
correlations between the Texas and North gardens, particularly in the
Gulf subpopulation and Both subpopulations (Figure~\ref{fig-map} B). If
these phenotypic correlations have a genetic component, they could be
partially or completely captured by the covariance structures specified
in the mash model.

@@

We used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood. Five
GxWeather covariance structures were selected, zero to two per
subpopulation: two for the start of vegetative growth, and three for
flowering date. Of these five, three had mass on them in mash models of
the strong effects (Figure~\ref{fig-covar} A-B). Two of these three
matrices had negative covariances between sites between Texas and North
gardens (Figure~\ref{fig-covar} A), while one had all positive or
near-zero covariances. SNP-associated phenotypic effects covaried with
different weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the mash model of Gulf vegetative growth fell on a
covariance matrix constructed using the covariance of daylength 14 days
prior to the date of vegetative growth. The covariance matrix for this
weather cue was very similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Mash
models of the timing of vegetative growth in the Midwest subpopulation
and Both subpopulations did not include this GxWeather covariance; the
Midwest had no weight on any GxWeather covariance, while Both
subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured
covariance in effect sizes for SNPs in the Gulf and Midwest
subpopulations. 33\% of SNP effects on flowering in the Gulf
subpopulation covaried with cumulative rainfall in the seven days prior
to flowering (Figure~\ref{fig-covar} D). 22.6\% of SNP effects on
flowering in the Midwest subpopulation covaried with day length change
in the two days prior to flowering (Figure~\ref{fig-covar} D), which had
negative covariances between Texas and North gardens. Neither of these
covariance matrices significantly improved the log-likelihood of the
\emph{mash} model of Both subpopulations, and no GxWeather covariance
matrices had non-zero mass in models of Both subpopulations
(Figure~\ref{fig-covar} E). In five of the six mash models of strong
effects, the GxWeather covariance matrices captured a minority of the
posterior weights of the strong effects (Figure~\ref{fig-covar} C,E); in
these five models, the majority of this mass was on various canonical
covariance matrices. These matrices included simple heterozygosity, with
intermediate, positive covariances between all gardens, and
garden-specific effects present only at one garden.

We next characterized the pairwise patterns of effects where we were
confident in the sign of the effect at both gardens. If the overall
pattern of phenotypic expression differs between a pair of gardens, this
difference may be comprised of SNP effects that differ in sign between
these gardens (effects with antagonistic pleiotropy), SNP effects that
differ in magnitude (effects that are large in one garden or region and
smaller in others), and SNP effects that are indistinguishable in the
two gardens (Figure~\ref{fig-effects} A). We used the local false sign
rate (lfsr), an analogue of the lfdr that establishes confidence in the
effect sign, not the effect's difference from zero, to determine
significance. We required lfsr significance (p \textless{} 0.05) in both
gardens to include effects. This means that our tests for antagonistic
pleiotropy, or a sign change between conditions, carry an equal
statistical burden to those for effects with the same sign.

@@

Identifying the environmental cues that are predictive of, or even
correlated with, plant phenotypic responses remains a major challenge to
studies interrogating gene action across many natural environments. The
GxWeather photoperiod and cumulative rainfall cues we identify here are
functions of the genotypes measured and capture only a minority of SNP
effects on flowering. We could only assign SNP effects to a GxWeather
covariance structures in four of the six phenotype \& genetic
subpopulations we modeled. It's likely we did not include some GxWeather
important to these phenological cues - for example, overwintering
parameters that might cause variation in the start of vegetative growth.
More generally, it is difficult to predict the time scales over which
individuals may integrate environmental cues, particularly in perennial
species which may integrate these cues over longer time scales. If this
integration time itself varies between individuals, the covariance
structures we modeled cannot reflect this, though these structures would
likely be highly correlated with GxWeather structures we did include.
Our approach offers an opportunity to specify multiple environmental
cues and compete them to explain patterns of genetic effects, allowing
us to detect how important these cues are genome-wide, and how strongly
each cue influences each SNP. This is a key development to further
improve our understanding of genetic variation in GxE.

@@

To evaluate the prevalence and kinds of covariance patterns of SNP
effects across our common gardens, we used multivariate adaptive
shrinkage (\emph{mash}) on SNP effect estimates from the diversity panel
(35). To do this, we first conducted univarate genome-wide association
(GWAS) on site-specific best linear unbiased predictors (BLUPs) for the
timing of vegetative growth and flowering, using the switchgrassGWAS
package and the methods in (40). We used the effect estimates for single
nucleotide polymorphisms (SNPs) from these GWAS as preliminary effect
estimates for the two traits at each garden. We then jointly modeled
genetic effect estimates, and the mixture of ways these effects might
covary, across all eight common gardens, using a two-part procedure. We
first modeled genetic effects with different combinations of covariance
structures, using the effect size estimates from a subset of \ul{19K}
relatively unlinked (r\textsuperscript{2} \textless{} 0.2), randomly
selected SNPs used in the GWAS. In these models (hereafter referred to
as ``random''), we then specified a variety of potential covariance
structures among genetic effect size estimates from each garden (as
expanded on below) and used a greedy algorithm to iteratively select and
add covariance structures that significantly improved the log likelihood
of the \emph{mash} model. Given the high correlation between some of the
covariance patterns we tested, this approach allowed us to select only
the subset of patterns of covariance that significantly improved the fit
of the jointly inferred structure of genetic effect estimates present in
the data. We then selected 19K relatively unlinked (r\textsuperscript{2}
\textless{} 0.2) SNPs from each univariate GWAS that had the largest
effect estimates in each GWAS (hereafter ``strong'' effects), which gave
sets of \ul{\textbf{48K - 80K}} SNPs, and jointly re-estimated these
effect size estimates at all eight sites using the set of significant
covariance matrices that most improved the model fit on the random
effects. Notably, the majority of these ``strong'' effects were not
significant in univariate GWAS.
